Locality and Effect Reflection

### 2025-07-06 paper.tex
Continue to write

\lstdefinestyle{ocaml}{
  keywordstyle=\color{blue},
  emphstyle=\color{violet},
  language=[Objective]Caml,
  morekeywords={effect, exclave},
  escapeinside={(*@}{@*)},
  literate=
    {[]}{$\glob$}{1}
}

@@

\newcommand{\nat}{\mathbb{N}}
\newcommand{\glob}{\mathop{\Box}}
\newcommand{\cat}{\mathbb{C}}
\newcommand{\op}{\mathtt{op}}
\newcommand{\catop}{\cat^{\op}}
\newcommand{\id}[1]{\mathrm{id}_{#1}}
\newcommand{\setc}{\mathrm{Set}}
\newcommand{\psh}{\widehat{\cat}}
\newcommand{\terminal}{\top}
\newcommand{\initial}{\bot}
\newcommand{\lawc}{\mathrm{Law}}
\newcommand{\lawcop}{\lawc^{\mathtt{op}}}
\newcommand{\pshlawcop}{\widehat{\lawcop}}
\newcommand{\freef}{Free}
\newcommand{\constfname}{\Omega}
\newcommand{\constf}[1]{\constfname #1}
\newcommand{\yonedaname}{y}
\newcommand{\yoneda}[1]{y(#1)}
\newcommand{\homset}[3]{\mathrm{Hom}_{#1}(#2, #3)}
\newcommand{\arrow}[2]{#1 \Rightarrow #2}
\newcommand{\sem}[1]{\llbracket #1 \rrbracket}
\newcommand{\reflectname}{\Uparrow}
\newcommand{\reflectraw}[1]{\mathop{\reflectname_{#1}}}
\newcommand{\reflect}[3]{\reflectraw{#1}(#2, #3)}
\newcommand{\reflectc}[2]{\mathop{\reflectname_{#1 #2}}}
\newcommand{\performraw}[2]{\mathop{\reflectname(#1(#2))}}
\newcommand{\perform}[5]{\performraw{#1}{#2}(#3, #4. #5)}
\newcommand{\reifyname}{\Downarrow}
\newcommand{\reifyraw}[1]{\mathop{\reifyname_{#1}}}
\newcommand{\reify}[3]{\reifyraw{#1}(#2.\,#3)}
\newcommand{\reifyc}[2]{\mathop{\reifyname_{#1 #2}}}
\newcommand{\outl}[1]{\mathop{\mathrm{out}_L} #1}
\newcommand{\outr}[1]{\mathop{\mathrm{out}_R} #1}
\newcommand{\mon}{\mathcal{T}}
\newcommand{\ind}[1]{W_{#1}}
\newcommand{\indintro}[4]{#2_{#1}(#3, #4)}
\newcommand{\retname}{\mathrm{ret}}
\newcommand{\ret}[2]{\mathop{\retname_{#1}}(#2)}
\newcommand{\indelim}[3]{\mathop{\mathrm{rec}_{#1}} #2 \{#3\}}
\newcommand{\indretcase}[2]{\retname(#1) \Rightarrow #2}
\newcommand{\indcase}[4]{#1(#2, #3) \Rightarrow #4}
\newcommand{\seq}{\,;\,}
\newcommand{\fst}{\mathop{\pi_1}}
\newcommand{\snd}{\mathop{\pi_2}}
\newcommand{\inl}[1]{\mathop{\mathrm{in}_L} #1}
\newcommand{\inr}[1]{\mathop{\mathrm{in}_R} #1}
\newcommand{\case}[5]{\mathop{\mathrm{case}} #1 \,\{\,\inl{#2} \Rightarrow #3 \seq \inr{#4} \Rightarrow #5 \,\}}
\newcommand{\abs}[3]{\mathop{\lambda}(#1 \types #2).\,#3}
\newcommand{\app}[2]{#1\,#2}
\newcommand{\types}{\mathrel{:}}
\newcommand{\gtypes}{\mathrel{:_{\Box}}}
\newcommand{\cempty}{\varepsilon}
\newcommand{\ccons}[2]{#1;\,#2}
\newcommand{\lbind}[3]{\ccons{#1}{#2\types#3}}
\newcommand{\gbind}[3]{\ccons{#1}{#2\gtypes#3}}
\newcommand{\cdiv}[1]{#1/\Box}
\newcommand{\return}[1]{\mathop{\mathrm{return}} #1}
\newcommand{\letv}[3]{\mathop{\mathrm{let}} #1 = #2 \mathop{\mathrm{in}} #3}
\newcommand{\gbox}[1]{\mathop{\mathrm{box}} #1}
\newcommand{\gunbox}[4]{\mathop{\mathrm{let}} (\gbox #1) = #2 \mathop{\mathrm{in}} #3}
\newcommand{\grun}[1]{\mathop{\mathrm{run}}(#1)}
\newcommand{\sbst}[3]{#1[#2\backslash#3]}
\newcommand{\turnv}{\mathrel{\vdash_V}}
\newcommand{\turnc}{\mathrel{\vdash_C}}
\newcommand{\evmap}{\mathop{ev}}
\newcommand{\currymap}{\mathop{\lambda}}
\newcommand{\reducesto}{\rightsquigarrow}

@@

For example, in OCaml, one can write:
\begin{lstlisting}[style=ocaml]
type 'a Effect.t +=
  | Get : int Effect.t
  | Set : int -> unit Effect.t

let handle_state (f : unit -> 'a) : 'a =
  match f () with
  | x -> fun _ -> x
  | effect Get, k -> fun s -> continue k s s
  | effect Set s', k -> fun _ -> continue k () s'
\end{lstlisting}
and for the duration of the \lstinline[style=ocaml]{f ()} function call
the ambient monad is extended with two additional effectful operations:
\begin{lstlisting}[style=ocaml]
let get () : int = perform Get
let set (i : int) : unit = perform (Set i)
\end{lstlisting}

Note that, in OCaml, nothing prevents attempting to perform an algebraic
effect operation that is not supported by the current ambient monad,
resulting in an \lstinline[style=ocaml]{Unhandled} exception being
raised. Other languages, e.g. Koka\cite{??}, use an effect system to
statically ensure operations are only performed when they are available.

@@

\paragraph{Effect reflection in OxCaml}
OxCaml\cite{lorenzen2024oxidizing} provides type system support for
tracking locality and runtime support for algebraic effects. For
backwards compatibility with OCaml, an unadorned arrow type
\lstinline[style=ocaml]{t -> s} is treated as if the parameter and
return types were actually \lstinline[style=ocaml]{[]t} and
\lstinline[style=ocaml]{[]s}. To get parameters or results that are not
implicitly global, we write
e.g. \lstinline[style=ocaml]{t @ local -> s @ local}.

Using this support, with can provide effect reflection as a simple API
that we can use for programming with effects.  For example, we can use
it to write an effect handler equivalent to the OCaml one above:
\begin{lstlisting}[style=ocaml]
module State = struct
  type 'a t =
    | Get : int t
    | Set : int -> unit t
end

@@

let handle_state (f : Handler(State).t @ local -> 'a) (i : int) : 'a =
  let rec loop s : 'a Term(State).t @ local -> 'a  = function
    | Return x -> x
    | Op(Get, k) -> loop s (k s)
    | Op(Set s', k) -> loop s' (k ())
  in
  loop i (State_eff.reify_local f)
\end{lstlisting}
which calls \lstinline[style=ocaml]{f} and passes it a local value of
type \lstinline[style=ocaml]{Handler(State).t}. For the duration of the
call to \lstinline[style=ocaml]{f} the ambient monad is extended with
two additional effectful operations, which can be used via reflection:
\begin{lstlisting}[style=ocaml]
let get (h : Handler(State).t @ local) : int =
  State_eff.perform Get h

@@

Values in OxCaml are associated with both a type -- which describe how
values can be introduced and eliminated -- and a \emph{mode} -- which
track properties related to operations other than introduction and
elimination. Modes are a product of a number of different axes, tracking
a variety of different properties, but here we are only interested in
one of them: \emph{locality}. Ignoring the other axes, a mode is either
\lstinline[style=ocaml]{local} or \lstinline[style=ocaml]{global}.

Locality tracks the ability of a value to leave \emph{regions}.  Regions
describe some period of time during a program's execution. Every
function call defines a corresponding region that, by default, lasts for
the entire extent of the call. This means that, by default,
\lstinline[style=ocaml]{local} values cannot be returned from a
function.
\begin{lstlisting}[style=ocaml]
let leak_local x =
  let (o @ local) = Some x in
  o
(*@ \textcolor{red}{Error: o} @*)
       (*@ \textcolor{red}{\textasciicircum} @*)
(*@ \textcolor{red}{This value is local and would escape its region.} @*)
\end{lstlisting}

To support functions that return local values, the region of a function
can be ended early using the \lstinline[style=ocaml]{exclave} construct.
\begin{lstlisting}[style=ocaml]
let return_local x =
  exclave
    (let (o @ local) = Some x in
     o)
\end{lstlisting}
\lstinline[style=ocaml]{exclave e} can only appear in tail position
within a function.  It ends the function's region early and then
executes the expression
\lstinline[style=ocaml]{e}. \lstinline[style=ocaml]{e} is essentially
executed within the region of the caller of the function and any local
values created within it can be safely returned to that caller.

The place where modes appear within the type algebra is on arrow
types. A function being \lstinline[style=ocaml]{global} doesn't tell you
that the values it accepts or returns should be
\lstinline[style=ocaml]{global} too, so instead the arrow type is
labelled with two modes: one for the argument and one for the return
value.
\begin{lstlisting}[style=ocaml]
val return_local : 'a @ local -> 'a option @ local
\end{lstlisting}
For convenience, and backwards compatibility with OCaml, either mode can
be omitted, in which case it defaults to
\lstinline[style=ocaml]{global}.

By default, modes are deep: a local \lstinline[style=ocaml]{string list}
is a local list of local strings. However, the depth that a mode applies
can be controlled via \emph{modalities}. Record fields can be annotated
with \lstinline[style=ocaml]{global} modality to indicate that the
contents of the field are global even when the surrounding record is
local. For example:
\begin{lstlisting}[style=ocaml]
type 'a gbl = { g : 'a @@ global }
\end{lstlisting}
gives a record whose only field has the \lstinline[style=ocaml]{global}
modality. The contents of the \lstinline[style=ocaml]{g} field is always
global, for example:
\begin{lstlisting}[style=ocaml]
let project_global (r @ local) =
  r.g
\end{lstlisting}
gives a function of type:
\begin{lstlisting}[style=ocaml]
val project_global : 'a gbl @ local -> 'a @ global
\end{lstlisting}
Variant constructor arguments can also be annotated with modalities.

Note that, whilst there is a \lstinline[style=ocaml]{global} modality,
there is no \lstinline[style=ocaml]{local} modality. It does not make
sense for a value which has the ability to leave a region to contain a
value which doesn't, as when the first value leaves a region it will
implicitly cause the second one to leave that region as well. This means
the \lstinline[style=ocaml]{global} mode is always deep.

@@

The types and operations of our effect reflection API are parameterized
by type constructors representing the operations of an algebraic
effect. In OxCaml, such higher-kinded interfaces must be built using
functors, so we start with a simple module type
\lstinline[style=ocaml]{Op} to act as the parameter type for these
functors:
\begin{lstlisting}[style=ocaml]
module type Op = sig type 'a t end
\end{lstlisting}
The fundamental types of the API are the \lstinline[style=ocaml]{Term}
and \lstinline[style=ocaml]{Handler} types.
\begin{lstlisting}[style=ocaml]
module Term (O : Op) : sig
  type 'a t =
    | Return : 'a @@ global -> 'a t
    | Op : 'r O.t @@ global * ('r -> 'a t) -> 'a t
end

module Handler(O : Op) : sig
  type t
end
\end{lstlisting}
For an effect signature \lstinline[style=ocaml]{O},
\lstinline[style=ocaml]{'a Term(O).t} is the type of terms in the free
algebra of the corresponding effect -- otherwise known as the freer
monad\cite{??}. \lstinline[style=ocaml]{Handler(O).t} is the type of
mappings from the operations of \lstinline[style=ocaml]{O} into the
operations of the current ambient monad. Alternatively,
\lstinline[style=ocaml]{Handler(O).t} can be described as representing
algebraic effect handlers that are handling the current computation --
hence the name.

The main piece of the API is a family of isomorphisms:
\begin{equation*}
  \text{\lstinline[style=ocaml]{'a Term(O).t}} \qquad
  \cong \qquad \text{\lstinline[style=ocaml]{Handler(O).t @ local -> 'a}}
\end{equation*}
Due to the absence of mode polymorphism in OxCaml, we expose this
isomorphism as three functions: \lstinline[style=ocaml]{reify} and
\lstinline[style=ocaml]{reify_local} which implement the left-to-right
direction at the \lstinline[style=ocaml]{global} and
\lstinline[style=ocaml]{local} modes respectively; and
\lstinline[style=ocaml]{perform}, which is equivalent to the
right-to-left direction but makes for a more convenient primitive in
practice. \lstinline[style=ocaml]{perform} amounts to reflection of
individual operations as opposed to full terms.
\begin{lstlisting}[style=ocaml]
module Reflection (O : Op) : sig
  val reify : (Handler(O).t @ local -> 'a) -> 'a Term(O).t

@@

  val perform : 'a O.t -> (Handler(O).t @ local -> 'a)
end
\end{lstlisting}
Using \lstinline[style=ocaml]{perform} we can implement reflection of
terms as:
\begin{lstlisting}[style=ocaml]
let rec reflect t h =
  match t with
  | Return x -> x
  | Op(op, k) -> reflect (k (perform op h)) h

let rec reflect_local (t @ local) h =
  match t with
  | Return x -> x
  | Op(op, k) -> reflect_local (k (perform op h)) h
\end{lstlisting}
with types:
\begin{lstlisting}[style=ocaml]
val reflect : 'a Term(O).t -> (Handler(O).t @ local -> 'a)

@@

As shown in the introduction we can implement an effect handler for
state using this API. We start by defining the operations of the effect:
\begin{lstlisting}[style=ocaml]
module State = struct
  type 'a t =
    | Get : int t
    | Set : int -> unit t
end
\end{lstlisting}
Then we instantiate the reflection isomorphism for that effect and use
it to write a handler:
\begin{lstlisting}[style=ocaml]
module State_eff = Reflection(State)

let handle_state f i =
  let rec loop s = function
    | Term(State).Return x -> x
    | Term(State).Op(Get, k) -> loop s (k s)
    | Term(State).Op(Set s', k) -> loop s' (k ())
  in
  loop i (State_eff.reify_local f)
\end{lstlisting}
Finally we add an alias for the handler type and some convenient
wrappers for the generic effects:
\begin{lstlisting}[style=ocaml]
type handler = Handler(State).t

@@

let set h i =
  State_eff.perform (Set i) h
\end{lstlisting}
Giving us the following interface:
\begin{lstlisting}[style=ocaml]
type handler

val handle_state : (handler @ local -> 'a) @ local -> int -> 'a
val get : handler @ local -> int
val set : handler @ local -> int -> unit
\end{lstlisting}
Note that we are able to use \lstinline[style=ocaml]{reify_local} to allow
the function passed to \lstinline[style=ocaml]{handle_state} to be local.

This shows how to implement state using effect reflection, but note that
the same interface can also be implemented direcly using mutable state
and locality:
\begin{lstlisting}[style=ocaml]
type handler = int ref

@@

The following function uses state to calculate the total of a list of
integers:
\begin{lstlisting}[style=ocaml]
let total l =
  handle_state (fun h ->
    List.iter (fun x -> set h (get h + x)) l;
    get h)
\end{lstlisting}
This relies on the locality of the function parameter of
\lstinline[style=ocaml]{List.iter}:
\begin{lstlisting}[style=ocaml]
val iter : ('a -> unit) @ local -> 'a list -> unit
\end{lstlisting}
The \lstinline[style=ocaml]{local} annotation on the parameter ensures
that any function passed will not escape the current region. This allows
us to pass in a closure that closes over the
\lstinline[style=ocaml]{local} handler \lstinline[style=ocaml]{h} and
use it to perform some effects.

This ability for higher-order functions like
\lstinline[style=ocaml]{List.iter} to be reused with different ambient
monads achieves the ``lightweight effect polymorphism'' described by ??
et al.\cite{??} without requiring a separate notion of second-class
function.

@@

A more realistic use of effect reflection is a \emph{generator}
effect. This effect has a single operation:
\begin{lstlisting}[style=ocaml]
module Gen = struct
  type 'a t =
    | Gen : int -> unit t
end
\end{lstlisting}
Along with a handler that turns computations using
\lstinline[style=ocaml]{Gen} into streams of integers.
\begin{lstlisting}[style=ocaml]
module Gen_eff = Reflection(Gen)

@@

Unlike the state handler, here we must use
\lstinline[style=ocaml]{reify} rather than
\lstinline[style=ocaml]{reify_local} because our intention is that the
resulting stream of integers be a global value that can escape the
current region. Since it closes over the continuation
\lstinline[style=ocaml]{k}, we must use \lstinline[style=ocaml]{reify}
and require that \lstinline[style=ocaml]{f} be global:
\begin{lstlisting}[style=ocaml]
val handle_gen : (Handler(Gen).t @ local -> unit) -> ints
\end{lstlisting}

@@

Perhaps the most common use case for algebraic effects is for
concurrency and asynchronous I/O. Implementing asynchronous I/O from
scratch is beyond the scope of this paper, but we can build a simple
version on top of an existing monadic concurrency library. Assuming we
have a monadic future type \lstinline[style=ocaml]{'a Deferred.t} we can
write concurrent code in direct style by using an effect with a single
\lstinline[style=ocaml]{Await} operation that waits for a future to be
completed:
\begin{lstlisting}[style=ocaml]
module Await : sig
  type 'a t =
    | Await : 'a Deferred.t -> 'a t
end
\end{lstlisting}

\begin{lstlisting}[style=ocaml]
module Await_eff = Reflection(Await)

@@

\lstinline[style=ocaml]{Sum(L)(R)} is the coproduct of the effects
\lstinline[style=ocaml]{L} and \lstinline[style=ocaml]{R}.

>>>>

\lstdefinestyle{oxcaml}{
  keywordstyle=\color{blue},
  emphstyle=\color{violet},
  language=[Objective]Caml,
  morekeywords={effect, exclave},
  escapeinside={(*@}{@*)},
  literate=
    {[]}{$\glob$}{1}
}

@@

\newcommand{\nat}{\mathbb{N}}
\newcommand{\glob}{\mathop{\Box}}
\newcommand{\cat}{\mathbb{C}}
\newcommand{\catobj}{C}
\newcommand{\op}{\mathtt{op}}
\newcommand{\catop}{\cat^{\op}}
\newcommand{\id}[1]{\mathrm{id}_{#1}}
\newcommand{\setc}{\mathrm{Set}}
\newcommand{\psh}{\widehat{\cat}}
\newcommand{\terminal}{\top}
\newcommand{\initial}{\bot}
\newcommand{\lawc}{\mathrm{Law}}
\newcommand{\lawcop}{\lawc^{\mathtt{op}}}
\newcommand{\pshlawcop}{\widehat{\lawcop}}
\newcommand{\freef}{Free}
\newcommand{\constfname}{\Omega}
\newcommand{\constf}[1]{\constfname #1}
\newcommand{\yonedaname}{y}
\newcommand{\yoneda}[1]{y(#1)}
\newcommand{\homset}[3]{\mathrm{Hom}_{#1}(#2, #3)}
\newcommand{\arrow}[2]{#1 \Rightarrow #2}
\newcommand{\sem}[1]{\llbracket #1 \rrbracket}
\newcommand{\reflectname}{\Uparrow}
\newcommand{\reflectraw}[1]{\mathop{\reflectname_{#1}}}
\newcommand{\reflect}[3]{\reflectraw{#1}(#2, #3)}
\newcommand{\reflectc}[2]{\mathop{\reflectname_{#1 #2}}}
\newcommand{\performraw}[2]{\mathop{\reflectname(#1(#2))}}
\newcommand{\perform}[5]{\performraw{#1}{#2}(#3, #4. #5)}
\newcommand{\reifyname}{\Downarrow}
\newcommand{\reifyraw}[1]{\mathop{\reifyname_{#1}}}
\newcommand{\reify}[3]{\reifyraw{#1}(#2.\,#3)}
\newcommand{\reifyc}[2]{\mathop{\reifyname_{#1 #2}}}
\newcommand{\outl}[1]{\mathop{\mathrm{out}_L} #1}
\newcommand{\outr}[1]{\mathop{\mathrm{out}_R} #1}
\newcommand{\mon}{\mathcal{T}}
\newcommand{\ind}[1]{W_{#1}}
\newcommand{\indintro}[4]{#2_{#1}(#3,\, #4)}
\newcommand{\retname}{\mathrm{ret}}
\newcommand{\ret}[2]{\mathop{\retname_{#1}}(#2)}
\newcommand{\indelim}[3]{\mathop{\mathrm{rec}_{#1}} #2 \,\{#3\}}
\newcommand{\indretcase}[2]{\retname(#1) \Rightarrow #2}
\newcommand{\indcase}[4]{#1(#2, #3) \Rightarrow #4}
\newcommand{\seq}{\,;\,}
\newcommand{\fst}{\mathop{\pi_1}}
\newcommand{\snd}{\mathop{\pi_2}}
\newcommand{\inl}[1]{\mathop{\mathrm{in}_L} #1}
\newcommand{\inr}[1]{\mathop{\mathrm{in}_R} #1}
\newcommand{\case}[5]{\mathop{\mathrm{case}} #1 \,\{\,\inl{#2} \Rightarrow #3 \seq \inr{#4} \Rightarrow #5 \,\}}
\newcommand{\abs}[3]{\mathop{\lambda}(#1 \types #2).\,#3}
\newcommand{\app}[2]{#1\,#2}
\newcommand{\types}{\mathrel{:}}
\newcommand{\gtypes}{\mathrel{:_{\Box}}}
\newcommand{\cempty}{\varepsilon}
\newcommand{\ccons}[2]{#1;\,#2}
\newcommand{\lbind}[3]{\ccons{#1}{#2\types#3}}
\newcommand{\gbind}[3]{\ccons{#1}{#2\gtypes#3}}
\newcommand{\cdiv}[1]{#1/\Box}
\newcommand{\return}[1]{\mathop{\mathrm{return}} #1}
\newcommand{\letv}[3]{\mathop{\mathrm{let}} #1 = #2 \mathop{\mathrm{in}} #3}
\newcommand{\gbox}[1]{\mathop{\mathrm{box}} #1}
\newcommand{\gunbox}[3]{\mathop{\mathrm{let}} (\gbox #1) = #2 \mathop{\mathrm{in}} #3}
\newcommand{\grun}[1]{\mathop{\mathrm{run}}(#1)}
\newcommand{\sbst}[3]{#1[#2\backslash#3]}
\newcommand{\turnv}{\mathrel{\vdash_V}}
\newcommand{\turnc}{\mathrel{\vdash_C}}
\newcommand{\evmap}{\mathop{ev}}
\newcommand{\currymap}{\mathop{\lambda}}
\newcommand{\reducestov}{\rightsquigarrow_V}
\newcommand{\reducestoc}{\rightsquigarrow_C}
\newcommand{\mif}{\mathop{\mathrm{if}}}
\newcommand{\mthen}{\mathop{\mathrm{then}}}
\newcommand{\melse}{\mathop{\mathrm{else}}}
\newcommand{\embed}[1]{\lbag #1 \rbag}

@@

For example, in OCaml, one can write:
\begin{lstlisting}[style=oxcaml]
type 'a Effect.t +=
  | Get : int Effect.t
  | Set : int -> unit Effect.t

let handle_state (f : unit -> 'a) : 'a =
  match f () with
  | x -> fun _ -> x
  | effect Get, k -> fun s -> continue k s s
  | effect Set s', k -> fun _ -> continue k () s'
\end{lstlisting}
and for the duration of the \lstinline[style=oxcaml]{f ()} function call
the ambient monad is extended with two additional effectful operations:
\begin{lstlisting}[style=oxcaml]
let get () : int = perform Get
let set (i : int) : unit = perform (Set i)
\end{lstlisting}

Note that, in OCaml, nothing prevents attempting to perform an algebraic
effect operation that is not supported by the current ambient monad,
resulting in an \lstinline[style=oxcaml]{Unhandled} exception being
raised. Other languages, e.g. Koka\cite{??}, use an effect system to
statically ensure operations are only performed when they are available.

@@

\paragraph{Effect reflection in OxCaml}
OxCaml\cite{lorenzen2024oxidizing} provides type system support for
tracking locality and runtime support for algebraic effects. For
backwards compatibility with OCaml, an unadorned arrow type
\lstinline[style=oxcaml]{t -> s} is treated as if the parameter and
return types were actually \lstinline[style=oxcaml]{[]t} and
\lstinline[style=oxcaml]{[]s}. To get parameters or results that are not
implicitly global, we write
e.g. \lstinline[style=oxcaml]{t @ local -> s @ local}.

Using this support, with can provide effect reflection as a simple API
that we can use for programming with effects.  For example, we can use
it to write an effect handler equivalent to the OCaml one above:
\begin{lstlisting}[style=oxcaml]
module State = struct
  type 'a t =
    | Get : int t
    | Set : int -> unit t
end

@@

let handle_state (f : Handler(State).t @ local -> 'a) (i : int) : 'a =
  let rec loop s : 'a Term(State).t @ local -> 'a  = function
    | Return x -> x
    | Op(Get, k) -> loop s (k s)
    | Op(Set s', k) -> loop s' (k ())
  in
  loop i (State_eff.reify_local f)
\end{lstlisting}
which calls \lstinline[style=oxcaml]{f} and passes it a local value of
type \lstinline[style=oxcaml]{Handler(State).t}. For the duration of the
call to \lstinline[style=oxcaml]{f} the ambient monad is extended with
two additional effectful operations, which can be used via reflection:
\begin{lstlisting}[style=oxcaml]
let get (h : Handler(State).t @ local) : int =
  State_eff.perform Get h

@@

Values in OxCaml are associated with both a type -- which describe how
values can be introduced and eliminated -- and a \emph{mode} -- which
track properties related to operations other than introduction and
elimination. Modes are a product of a number of different axes, tracking
a variety of different properties, but here we are only interested in
one of them: \emph{locality}. Ignoring the other axes, a mode is either
\lstinline[style=oxcaml]{local} or \lstinline[style=oxcaml]{global}.

Locality tracks the ability of a value to leave \emph{regions}.  Regions
describe some period of time during a program's execution. Every
function call defines a corresponding region that, by default, lasts for
the entire extent of the call. This means that, by default,
\lstinline[style=oxcaml]{local} values cannot be returned from a
function.
\begin{lstlisting}[style=oxcaml]
let leak_local x =
  let (o @ local) = Some x in
  o
(*@ \textcolor{red}{Error: o} @*)
       (*@ \textcolor{red}{\textasciicircum} @*)
(*@ \textcolor{red}{This value is local and would escape its region.} @*)
\end{lstlisting}

To support functions that return local values, the region of a function
can be ended early using the \lstinline[style=oxcaml]{exclave} construct.
\begin{lstlisting}[style=oxcaml]
let return_local x =
  exclave
    (let (o @ local) = Some x in
     o)
\end{lstlisting}
\lstinline[style=oxcaml]{exclave e} can only appear in tail position
within a function.  It ends the function's region early and then
executes the expression
\lstinline[style=oxcaml]{e}. \lstinline[style=oxcaml]{e} is essentially
executed within the region of the caller of the function and any local
values created within it can be safely returned to that caller.

The place where modes appear within the type algebra is on arrow
types. A function being \lstinline[style=oxcaml]{global} doesn't tell you
that the values it accepts or returns should be
\lstinline[style=oxcaml]{global} too, so instead the arrow type is
labelled with two modes: one for the argument and one for the return
value.
\begin{lstlisting}[style=oxcaml]
val return_local : 'a @ local -> 'a option @ local
\end{lstlisting}
For convenience, and backwards compatibility with OCaml, either mode can
be omitted, in which case it defaults to
\lstinline[style=oxcaml]{global}.

By default, modes are deep: a local \lstinline[style=oxcaml]{string list}
is a local list of local strings. However, the depth that a mode applies
can be controlled via \emph{modalities}. Record fields can be annotated
with \lstinline[style=oxcaml]{global} modality to indicate that the
contents of the field are global even when the surrounding record is
local. For example:
\begin{lstlisting}[style=oxcaml]
type 'a gbl = { g : 'a @@ global }
\end{lstlisting}
gives a record whose only field has the \lstinline[style=oxcaml]{global}
modality. The contents of the \lstinline[style=oxcaml]{g} field is always
global, for example:
\begin{lstlisting}[style=oxcaml]
let project_global (r @ local) =
  r.g
\end{lstlisting}
gives a function of type:
\begin{lstlisting}[style=oxcaml]
val project_global : 'a gbl @ local -> 'a @ global
\end{lstlisting}
Variant constructor arguments can also be annotated with modalities.

Note that, whilst there is a \lstinline[style=oxcaml]{global} modality,
there is no \lstinline[style=oxcaml]{local} modality. It does not make
sense for a value which has the ability to leave a region to contain a
value which doesn't, as when the first value leaves a region it will
implicitly cause the second one to leave that region as well. This means
the \lstinline[style=oxcaml]{global} mode is always deep.

@@

The types and operations of our effect reflection API are parameterized
by type constructors representing the operations of an algebraic
effect. In OxCaml, such higher-kinded interfaces must be built using
functors, so we start with a simple module type
\lstinline[style=oxcaml]{Op} to act as the parameter type for these
functors:
\begin{lstlisting}[style=oxcaml]
module type Op = sig type 'a t end
\end{lstlisting}
The fundamental types of the API are the \lstinline[style=oxcaml]{Term}
and \lstinline[style=oxcaml]{Handler} types.
\begin{lstlisting}[style=oxcaml]
module Term (O : Op) : sig
  type 'a t =
    | Return : 'a @@ global -> 'a t
    | Op : 'r O.t @@ global * ('r -> 'a t) -> 'a t
end

module Handler(O : Op) : sig
  type t
end
\end{lstlisting}
For an effect signature \lstinline[style=oxcaml]{O},
\lstinline[style=oxcaml]{'a Term(O).t} is the type of terms in the free
algebra of the corresponding effect -- otherwise known as the freer
monad\cite{??}. \lstinline[style=oxcaml]{Handler(O).t} is the type of
mappings from the operations of \lstinline[style=oxcaml]{O} into the
operations of the current ambient monad. Alternatively,
\lstinline[style=oxcaml]{Handler(O).t} can be described as representing
algebraic effect handlers that are handling the current computation --
hence the name.

The main piece of the API is a family of isomorphisms:
\begin{equation*}
  \text{\lstinline[style=oxcaml]{'a Term(O).t}} \qquad
  \cong \qquad \text{\lstinline[style=oxcaml]{Handler(O).t @ local -> 'a}}
\end{equation*}
Due to the absence of mode polymorphism in OxCaml, we expose this
isomorphism as three functions: \lstinline[style=oxcaml]{reify} and
\lstinline[style=oxcaml]{reify_local} which implement the left-to-right
direction at the \lstinline[style=oxcaml]{global} and
\lstinline[style=oxcaml]{local} modes respectively; and
\lstinline[style=oxcaml]{perform}, which is equivalent to the
right-to-left direction but makes for a more convenient primitive in
practice. \lstinline[style=oxcaml]{perform} amounts to reflection of
individual operations as opposed to full terms.
\begin{lstlisting}[style=oxcaml]
module Reflection (O : Op) : sig
  val reify : (Handler(O).t @ local -> 'a) -> 'a Term(O).t

@@

  val perform : 'a O.t -> (Handler(O).t @ local -> 'a)
end
\end{lstlisting}
Using \lstinline[style=oxcaml]{perform} we can implement reflection of
terms as:
\begin{lstlisting}[style=oxcaml]
let rec reflect t h =
  match t with
  | Return x -> x
  | Op(op, k) -> reflect (k (perform op h)) h

let rec reflect_local (t @ local) h =
  match t with
  | Return x -> x
  | Op(op, k) -> reflect_local (k (perform op h)) h
\end{lstlisting}
with types:
\begin{lstlisting}[style=oxcaml]
val reflect : 'a Term(O).t -> (Handler(O).t @ local -> 'a)

@@

As shown in the introduction we can implement an effect handler for
state using this API. We start by defining the operations of the effect:
\begin{lstlisting}[style=oxcaml]
module State = struct
  type 'a t =
    | Get : int t
    | Set : int -> unit t
end
\end{lstlisting}
Then we instantiate the reflection isomorphism for that effect and use
it to write a handler:
\begin{lstlisting}[style=oxcaml]
module State_eff = Reflection(State)

let handle_state f i =
  let rec loop s = function
    | Term(State).Return x -> x
    | Term(State).Op(Get, k) -> loop s (k s)
    | Term(State).Op(Set s', k) -> loop s' (k ())
  in
  loop i (State_eff.reify_local f)
\end{lstlisting}
Finally we add an alias for the handler type and some convenient
wrappers for the generic effects:
\begin{lstlisting}[style=oxcaml]
type handler = Handler(State).t

@@

let set h i =
  State_eff.perform (Set i) h
\end{lstlisting}
Giving us the following interface:
\begin{lstlisting}[style=oxcaml]
type handler

val handle_state : (handler @ local -> 'a) @ local -> int -> 'a
val get : handler @ local -> int
val set : handler @ local -> int -> unit
\end{lstlisting}
Note that we are able to use \lstinline[style=oxcaml]{reify_local} to allow
the function passed to \lstinline[style=oxcaml]{handle_state} to be local.

This shows how to implement state using effect reflection, but note that
the same interface can also be implemented direcly using mutable state
and locality:
\begin{lstlisting}[style=oxcaml]
type handler = int ref

@@

The following function uses state to calculate the total of a list of
integers:
\begin{lstlisting}[style=oxcaml]
let total l =
  handle_state (fun h ->
    List.iter (fun x -> set h (get h + x)) l;
    get h)
\end{lstlisting}
This relies on the locality of the function parameter of
\lstinline[style=oxcaml]{List.iter}:
\begin{lstlisting}[style=oxcaml]
val iter : ('a -> unit) @ local -> 'a list -> unit
\end{lstlisting}
The \lstinline[style=oxcaml]{local} annotation on the parameter ensures
that any function passed will not escape the current region. This allows
us to pass in a closure that closes over the
\lstinline[style=oxcaml]{local} handler \lstinline[style=oxcaml]{h} and
use it to perform some effects.

This ability for higher-order functions like
\lstinline[style=oxcaml]{List.iter} to be reused with different ambient
monads achieves the ``lightweight effect polymorphism'' described by ??
et al.\cite{??} without requiring a separate notion of second-class
function.

@@

A more realistic use of effect reflection is a \emph{generator}
effect. This effect has a single operation:
\begin{lstlisting}[style=oxcaml]
module Gen = struct
  type 'a t =
    | Gen : int -> unit t
end
\end{lstlisting}
Along with a handler that turns computations using
\lstinline[style=oxcaml]{Gen} into streams of integers.
\begin{lstlisting}[style=oxcaml]
module Gen_eff = Reflection(Gen)

@@

Unlike the state handler, here we must use
\lstinline[style=oxcaml]{reify} rather than
\lstinline[style=oxcaml]{reify_local} because our intention is that the
resulting stream of integers be a global value that can escape the
current region. Since it closes over the continuation
\lstinline[style=oxcaml]{k}, we must use \lstinline[style=oxcaml]{reify}
and require that \lstinline[style=oxcaml]{f} be global:
\begin{lstlisting}[style=oxcaml]
val handle_gen : (Handler(Gen).t @ local -> unit) -> ints
\end{lstlisting}

@@

Perhaps the most common use case for algebraic effects is for
concurrency and asynchronous I/O. Implementing asynchronous I/O from
scratch is beyond the scope of this paper, but we can build a simple
version on top of an existing monadic concurrency library. Assuming we
have a monadic future type \lstinline[style=oxcaml]{'a Deferred.t} we can
write concurrent code in direct style by using an effect with a single
\lstinline[style=oxcaml]{Await} operation that waits for a future to be
completed:
\begin{lstlisting}[style=oxcaml]
module Await : sig
  type 'a t =
    | Await : 'a Deferred.t -> 'a t
end
\end{lstlisting}

\begin{lstlisting}[style=oxcaml]
module Await_eff = Reflection(Await)

@@

\lstinline[style=oxcaml]{Sum(L)(R)} is the coproduct of the effects
\lstinline[style=oxcaml]{L} and \lstinline[style=oxcaml]{R}.
